The CrypTool Script: \\ Cryptography, Mathematics and more \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free eLearning program CrypTool\\ (9th edition -- distributed with CrypTool version 1.4.20)} \vspace{60pt}

### 2009-06-30 script/en/chapters/primes.tex
1.4.20 updates
Use the book class for the script, instead of the article class

\newpage
\hypertarget{Kapitel_2}{}
\section{Prime Numbers}
\label{Label_Kapitel_2}
(Bernhard Esslinger, May 1999; Updates Nov. 2000, Dec. 2001, June 2003, May 2005, March 2006, June 2007)

@@

% --------------------------------------------------------------------------
\subsection{What are prime numbers?}
\index{Prime number} \index{Number!prime}
Prime numbers are whole, positive numbers greater than or equal to $2$ that can
only be divided by 1 and themselves. All other natural numbers greater than or
equal to $2$ can be formed by multiplying prime numbers.

@@

% --------------------------------------------------------------------------
\subsection{Prime numbers in mathematics}\label{primesinmath}

@@

% --------------------------------------------------------------------------
\subsection{How many prime numbers are there?}

@@

% --------------------------------------------------------------------------
% --------------------------------------------------------------------------
\vskip + 20pt
\subsection{The search for extremely large primes}
\label{search_for_very_big_primes}   % chap. 3.4

@@

% --------------------------------------------------------------------------
\hypertarget{RecordPrimes}{}
\subsubsection{The 20 largest known primes (as of May 2007)}  % Eyecatcher_neue-Mersenne
\label{RecordPrimes}
\index{Prime number!records}

@@

% --------------------------------------------------------------------------
\hypertarget{MersenneNumbers01}{}
\subsubsection{Special number types -- Mersenne numbers and Mersenne primes} 
\label{zahlentyp_mersenne}
\index{Mersenne!number}

@@

% --------------------------------------------------------------------------
\vskip +25 pt
\subsubsection{Challenge of the Electronic Frontier Foundation (EFF)}\index{EFF}
This search is also spurred on by a competition started by the non-profit
organisation EFF (Electronic Frontier Foundation) using the means of an 
unknown donator. The participants are rewarded with a total of 500,000 USD if
they find the longest prime number. In promoting this project, the unknown
donator is not looking for the quickest computer, but rather wants to draw 
people's attention to the opportunities offered by {\em cooperative networking} \\
{\href{http://www.eff.org/coopawards/prime-release1.html}{\tt http://www.eff.org/coopawards/prime-release1.html}}

@@

% --------------------------------------------------------------------------
\subsection[Prime number tests]{Prime number tests\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can apply the
    tests of Fermat and of Miller-Rabin:
    See learning units 3.2 and 3.3, pages 3-11/11.\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix \ref{s:appendix-Learn-NT}.
}
\label{primality_tests}   % chap. 3.5
\index{Prime number!test}

@@

% --------------------------------------------------------------------------
\vskip +30 pt
\subsection{Overview special number types and the search for a formula for primes}\label{spezialzahlentypen}
\index{Prime number!formula}
There are currently no useful, open (i.e. not recursive) formulae known that
only deliver prime numbers (recursive means that in order to calculate the
function the same function is used with a smaller variable). Mathematicians
would be happy if they could find a formula that leaves gaps (i.e. does not
deliver all prime numbers) but does not deliver any composite (non-prime)
numbers.

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{MersenneNumbers02}{}
\subsubsection{Mersenne numbers $f(n) = 2^n - 1$  \quad for $ n $ prime}
    \index{Prime number!Mersenne} \index{Mersenne!prime number}
    As shown \hyperlink{MersenneNumbers01}{above}, this formula seems
    to deliver relatively large prime numbers but - as for $n=11$ 
    [$f(n)=2,047$] - it is repeatedly the case that the result even with
    prime exponents is \hyperlink{Mer-nums-not-always-prim}{not} prime.\\
    Today, all the Mersenne primes having less than around 4,000,000 digits
    are known (\hyperlink{M-39}{M-39}\index{Mersenne!prime number!M-39}):
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}
\end{itemize}      % be_2005_UPDATEN_if-new-mersenne-prime-appears

% --------------------------------------------------------------------------
\vskip +10 pt
\subsubsection
   [Generalized Mersenne numbers $f(k,n) = k \cdot 2^n \pm 1$ / Proth numbers]
   {Generalized Mersenne numbers
   $f(k,n) = k \cdot 2^n \pm 1 $  for $ n $ prime and $ k $ small prime
    / Proth numbers\footnotemark}
    \footnotetext{%
       Their names come from the French farmer FranÁois Proth (1852-1879).
       More famous as the Proth primes is the related Sierpinski 
       problem\index{Number!Sierpinski}: Find all numbers k, so that
       $k*2^n+1$ is composite for all n > 0.
       See table \ref{L_n_Largest_Known-Primes}.
    }
   \label{generalized-mersenne-no1}

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{generalizedMersennenumbers}{}
\subsubsection
   [Generalized Mersenne numbers $ f(b,n) = b^n \pm 1 $ / Cunningham project]
   {Generalized Mersenne numbers 
   $ f(b,n) = b^n \pm 1 $ / The Cunningham project}

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{FermatNumbers02}{}
\subsubsection[Fermat numbers $f(n) = 2^{2^n} + 1$]
              {Fermat numbers\footnotemark~$f(n) = 2^{2^n} + 1$} 
    \footnotetext{%
       The Fermat prime numbers play a role in circle division.
       As proven by Gauss\index{Gauss, Carl Friedrich} a regular $p$-edge
       can only be constructed with the use of a pair of compasses and a 
       ruler, when $p$ is a Fermat prime number.
    }
    \label{L-FermatNumbers02}
    \index{Prime number!Fermat} \index{Fermat!prime number}       
    As mentioned \hyperlink{FermatNumbers01}{above} in 
    chapter~\ref{FermatNumbers01}, Fermat wrote to Mersenne regarding 
    his assumption, that all numbers of this type are primes.
    This assumption was disproved by Euler (1732). The prime $641$ divides $f(5)$\footnote{%
    Surprisingly this number can easily be found by using Fermat's theorem (see e.g. \cite[p. 176]{Scheid1994})
    }. 
%    Surprisingly he would have been able obtain a positive result using
%    the negative prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
f(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
f(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
f(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
f(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
f(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = \mbox{65,537} &  \mapsto {\rm ~prime}  \\
f(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = \mbox{4,294,967,297} = 641 \cdot \mbox{6,700,417} &  \mapsto {\rm ~NOT~prime} ! \\
f(6) = 2^{2^6} + 1  = 2^{64} + 1 &  = \mbox{18,446,744,073,709,551,617} \\
                                 &  = \mbox{274,177} \cdot \mbox{67,280,421,310,721} & \mapsto {\rm ~NOT~prime} !\\
f(7) = 2^{2^7} + 1  = 2^{128} + 1 & = \mbox{(see page~\pageref{F7Morrison})}  
																 & \mapsto {\rm ~NOT~prime} !

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{generalizedFermatprimes}{}
\subsubsection[Generalized Fermat numbers $f(b,n) = b^{2^n} + 1$]
              {Generalized Fermat numbers\footnotemark~$f(b,n) = b^{2^n} + 1$}
    \footnotetext{%
      The base of this power is no longer restricted to 2.
      Even more generic would be:  $f(b,c,n) = b^{c^n} \pm 1$
    }
\label{generalized-fermat}
\index{Fermat!number!generalized} 
    Generalized Fermat numbers are more numerous than Mersenne numbers of a
    equal size and many of them are waiting to be discovered to fill the
    big gaps between the Mersenne primes already found or still undiscovered.
    Progress in number theory made it possible that numbers, where the
    representation is not limited to the base 2, can be tested at almost
    the same speed than a Mersenne number.

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\subsubsection{Carmichael numbers\index{Number!Carmichael}}
As mentioned \hyperlink{HT-Carmichael-number01}{above} in
chapter~\ref{L-Carmichael-number01} not all Carmichael numbers are prime.

% --------------------------------------------------------------------------
\vskip +10 pt
\subsubsection{Pseudo prime numbers%
\index{Prime number!pseudo prime}\index{Number!pseudo prime}%
}
See \hyperlink{HT-Pseudoprimenumber01}{above} in
chapter~\ref{L-Pseudoprimenumber01}.

% --------------------------------------------------------------------------
\vskip +10 pt
\subsubsection{Strong pseudo prime numbers%
\index{Prime number!strong pseudo prime}\index{Number!strong pseudo prime}%
}
See \hyperlink{HT-Strongpseudoprimenumber01}{above} in
chapter~\ref{L-Strongpseudoprimenumber01}.

% --------------------------------------------------------------------------
\vskip +10 pt
\subsubsection
    [Idea based on Euclid's proof $p_1 \cdot p_2 \cdots p_n +1$]
    {Idea based on Euclid's proof $p_1 \cdot p_2 \cdots p_n +1$}
This idea is based on \hyperlink{euclid}{Euclid's proof} that there are
infinite many prime numbers.
$$
\begin{array}{lll}
2{\cdot}3 +1 &      = 7 &          \mapsto {\rm ~prime} \\
2{\cdot}3{\cdot}5 +1 &      = 31    &      \mapsto {\rm ~prime} \\
2{\cdot}3{\cdot}5{\cdot}7 +1 &      = 211   &      \mapsto {\rm ~prime} \\
2{\cdot}3{\cdots}11 +1 &        = 2,311  &      \mapsto {\rm ~prime} \\
2\cdot3 \cdots 13 +1 &  = 59 \cdot 509 &    \mapsto {\rm ~NOT~prime} ! \\
2\cdot3 \cdots 17 +1 &  = 19 \cdot 97 \cdot 277 &   \mapsto {\rm ~NOT~prime} !
\\
\end{array}
$$

% --------------------------------------------------------------------------
\vskip +10 pt
\subsubsection{As above but $-1$ except $+1$: $p_1 \cdot p_2 \cdots p_n -1$}
$$
\begin{array}{lll}
2\cdot 3 -1     &   = 5 &   \mapsto {\rm ~prime} \\
2\cdot 3 \cdot  5  -1   &   = 29 &  \mapsto {\rm ~prime} \\
2\cdot 3 \cdots 7  -1   &   = 11 \cdot 19 & \mapsto {\rm ~NOT~prime} ! \\
2\cdot 3 \cdots 11 -1   &   = 2,309 &    \mapsto {\rm ~prime} \\
2\cdot 3 \cdots 13 -1   &   = 30,029 &    \mapsto {\rm ~prime} \\
2\cdot 3 \cdots 17 -1    &  = 61 \cdot 8,369 &   \mapsto {\rm ~NOT~prime!}
\end{array} 
$$

% --------------------------------------------------------------------------
\vskip +10 pt
\subsubsection[Euclidean numbers $e_n = e_0 \cdot e_1 \cdots e_{n-1} + 1$]
              {Euclidean numbers $e_n = e_0 \cdot e_1 \cdots e_{n-1} + 1$
	       with $ n \geq 1 $ and $ e_0 := 1 $}
%	       with $n$ greater than or equal to $1$ and $e_0 := 1$}
    \index{Euclidean number}
    $e_{n-1}$ is not the $(n-1)$th prime number, but the number 
    previously found here.
    Unfortunately this formula is not open but recursive.
    The sequence starts with

@@

% --------------------------------------------------------------------------
% \pagebreak %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vskip +10 pt
\subsubsection{$f(n) = n^2 + n + 41$}
\label{L-Polynomfunktion01-41}
   This sequence starts off very {\em promisingly},  
   but is far from being a proof.

@@

% --------------------------------------------------------------------------
\pagebreak %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \vskip +10 pt
\subsubsection{$f(n) = n^2 - 79 \cdot n + 1,601$}
\label{L-Polynomfunktion02-1601}   \label{SrcPrimArith1}
   This function\footnote{%
   See \hyperlink{PrimArith1}{Appendix A of this chapter} for the source code
   to compute the table using Pari-GP.\index{Pari-GP}
}
   delivers prime numbers for all values from $n=0$ to $n=79$.
Unfortunately $f(80) = 1,681 = 11 \cdot 151$ is not a prime number. To this
date, no function has been found that delivers more prime numbers in a row. On
the other hand, each prime occurs twice (first in the decreasing then in the
increasing sequence), which means that the algorithm delivers a total of 40
different prime values (these are the same ones as delivered by the function
in chapter~\ref{L-Polynomfunktion01-41})\footnote{%
   Another quadratic polynom, which delivers these primes, is: 
   $f(n) = n^2 - 9 \cdot n + 61$.\\
   Among the first 1000 sequence elements more than 50\% are prime
   (See \hyperlink{PrimArith2}{Appendix A of this chapter}).}.
$$
\begin{array}{|ll||ll|}
\hline
f(0)  = 1.601 & \mapsto {\rm ~prim}   &   f(26) =   223 & \mapsto {\rm ~prim} \\
f(1)  = 1.523 & \mapsto {\rm ~prim}   &   f(27) =   197 & \mapsto {\rm ~prim} \\
f(2)  = 1.447 & \mapsto {\rm ~prim}   &   f(28) =   173 & \mapsto {\rm ~prim} \\
f(3)  = 1.373 & \mapsto {\rm ~prim}   &   f(29) =   151 & \mapsto {\rm ~prim} \\
f(4)  = 1.301 & \mapsto {\rm ~prim}   &   f(30) =   131 & \mapsto {\rm ~prim} \\
f(5)  = 1.231 & \mapsto {\rm ~prim}   &   f(31) =   113 & \mapsto {\rm ~prim} \\
f(6)  = 1.163 & \mapsto {\rm ~prim}   &   f(32) =    97 & \mapsto {\rm ~prim} \\
f(7)  = 1.097 & \mapsto {\rm ~prim}   &   f(33) =    83 & \mapsto {\rm ~prim} \\
f(8)  = 1.033 & \mapsto {\rm ~prim}   &   f(34) =    71 & \mapsto {\rm ~prim} \\
f(9)  =   971 & \mapsto {\rm ~prim}   &   f(35) =    61 & \mapsto {\rm ~prim} \\
f(10) =   911 & \mapsto {\rm ~prim}   &   f(36) =    53 & \mapsto {\rm ~prim} \\
f(11) =   853 & \mapsto {\rm ~prim}   &   f(37) =    47 & \mapsto {\rm ~prim} \\
f(12) =   797 & \mapsto {\rm ~prim}   &   f(38) =    43 & \mapsto {\rm ~prim} \\
f(13) =   743 & \mapsto {\rm ~prim}   &   f(39) =    41 & \mapsto {\rm ~prim} \\
f(14) =   691 & \mapsto {\rm ~prim}   &   f(40) =    41 & \mapsto {\rm ~prim} \\              
f(15) =   641 & \mapsto {\rm ~prim}   &   f(41) =    43 & \mapsto {\rm ~prim} \\              
f(16) =   593 & \mapsto {\rm ~prim}   &   f(42) =    47 & \mapsto {\rm ~prim} \\              
f(17) =   547 & \mapsto {\rm ~prim}   &   f(43) =    53 & \mapsto {\rm ~prim} \\              
f(18) =   503 & \mapsto {\rm ~prim}   &   \cdots  &  \\                                       
f(19) =   461 & \mapsto {\rm ~prim}   &   f(77) = 1.447 & \mapsto {\rm ~prim} \\             
f(20) =   421 & \mapsto {\rm ~prim}   &   f(78) = 1.523 & \mapsto {\rm ~prim} \\             
f(21) =   383 & \mapsto {\rm ~prim}   &   f(79) = 1.601 & \mapsto {\rm ~prim} \\             
f(22) =   347 & \mapsto {\rm ~prim}   &   f(80) = 41 \cdot 41 & \mapsto {\rm ~NOT~prim!} \\
f(21) =   383 & \mapsto {\rm ~prim}   &   f(81) = 41 \cdot 43 & \mapsto {\rm ~NOT~prim!} \\ 
f(22) =   347 & \mapsto {\rm ~prim}   &   f(82) = 1.847 & \mapsto {\rm ~prim} \\             
f(23) =   313 & \mapsto {\rm ~prim}   &   f(83) = 1.933 & \mapsto {\rm ~prim} \\              
f(24) =   281 & \mapsto {\rm ~prim}   &   f(84) = 43 \cdot 47 &  \mapsto {\rm ~NOT~prim!} \\
f(25) =   251 & \mapsto {\rm ~prim}   &   & \\
\hline
\end{array} 
$$

>>>>

\newpage
\hypertarget{Kapitel_2}{}
\chapter{Prime Numbers}
\label{Label_Kapitel_2}
(Bernhard Esslinger, May 1999; Updates Nov. 2000, Dec. 2001, June 2003, May 2005, March 2006, June 2007)

@@

% --------------------------------------------------------------------------
\section{What are prime numbers?}
\index{Prime number} \index{Number!prime}
Prime numbers are whole, positive numbers greater than or equal to $2$ that can
only be divided by 1 and themselves. All other natural numbers greater than or
equal to $2$ can be formed by multiplying prime numbers.

@@

% --------------------------------------------------------------------------
\section{Prime numbers in mathematics}\label{primesinmath}

@@

% --------------------------------------------------------------------------
\section{How many prime numbers are there?}

@@

% --------------------------------------------------------------------------
% --------------------------------------------------------------------------
\vskip + 20pt
\section{The search for extremely large primes}
\label{search_for_very_big_primes}   % chap. 3.4

@@

% --------------------------------------------------------------------------
\hypertarget{RecordPrimes}{}
\subsection{The 20 largest known primes (as of May 2007)}  % Eyecatcher_neue-Mersenne
\label{RecordPrimes}
\index{Prime number!records}

@@

% --------------------------------------------------------------------------
\hypertarget{MersenneNumbers01}{}
\subsection{Special number types -- Mersenne numbers and Mersenne primes} 
\label{zahlentyp_mersenne}
\index{Mersenne!number}

@@

% --------------------------------------------------------------------------
\vskip +25 pt
\subsection{Challenge of the Electronic Frontier Foundation (EFF)}\index{EFF}
This search is also spurred on by a competition started by the non-profit
organisation EFF (Electronic Frontier Foundation) using the means of an 
unknown donator. The participants are rewarded with a total of 500,000 USD if
they find the longest prime number. In promoting this project, the unknown
donator is not looking for the quickest computer, but rather wants to draw 
people's attention to the opportunities offered by {\em cooperative networking} \\
{\href{http://www.eff.org/coopawards/prime-release1.html}{\tt http://www.eff.org/coopawards/prime-release1.html}}

@@

% --------------------------------------------------------------------------
\section[Prime number tests]{Prime number tests\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can apply the
    tests of Fermat and of Miller-Rabin:
    See learning units 3.2 and 3.3, pages 3-11/11.\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix \ref{s:appendix-Learn-NT}.
}
\label{primality_tests}   % chap. 3.5
\index{Prime number!test}

@@

% --------------------------------------------------------------------------
\vskip +30 pt
\section{Overview special number types and the search for a formula for primes}\label{spezialzahlentypen}
\index{Prime number!formula}
There are currently no useful, open (i.e. not recursive) formulae known that
only deliver prime numbers (recursive means that in order to calculate the
function the same function is used with a smaller variable). Mathematicians
would be happy if they could find a formula that leaves gaps (i.e. does not
deliver all prime numbers) but does not deliver any composite (non-prime)
numbers.

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{MersenneNumbers02}{}
\subsection{Mersenne numbers $f(n) = 2^n - 1$  \quad for $ n $ prime}
    \index{Prime number!Mersenne} \index{Mersenne!prime number}
    As shown \hyperlink{MersenneNumbers01}{above}, this formula seems
    to deliver relatively large prime numbers but - as for $n=11$ 
    [$f(n)=2,047$] - it is repeatedly the case that the result even with
    prime exponents is \hyperlink{Mer-nums-not-always-prim}{not} prime.\\
    Today, all the Mersenne primes having less than around 4,000,000 digits
    are known (\hyperlink{M-39}{M-39}\index{Mersenne!prime number!M-39}):
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}
\end{itemize}      % be_2005_UPDATEN_if-new-mersenne-prime-appears

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection
   [Generalized Mersenne numbers $f(k,n) = k \cdot 2^n \pm 1$ / Proth numbers]
   {Generalized Mersenne numbers
   $f(k,n) = k \cdot 2^n \pm 1 $  for $ n $ prime and $ k $ small prime
    / Proth numbers\footnotemark}
    \footnotetext{%
       Their names come from the French farmer FranÁois Proth (1852-1879).
       More famous as the Proth primes is the related Sierpinski 
       problem\index{Number!Sierpinski}: Find all numbers k, so that
       $k*2^n+1$ is composite for all n > 0.
       See table \ref{L_n_Largest_Known-Primes}.
    }
   \label{generalized-mersenne-no1}

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{generalizedMersennenumbers}{}
\subsection
   [Generalized Mersenne numbers $ f(b,n) = b^n \pm 1 $ / Cunningham project]
   {Generalized Mersenne numbers 
   $ f(b,n) = b^n \pm 1 $ / The Cunningham project}

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{FermatNumbers02}{}
\subsection[Fermat numbers $f(n) = 2^{2^n} + 1$]
              {Fermat numbers\footnotemark~$f(n) = 2^{2^n} + 1$} 
    \footnotetext{%
       The Fermat prime numbers play a role in circle division.
       As proven by Gauss\index{Gauss, Carl Friedrich} a regular $p$-edge
       can only be constructed with the use of a pair of compasses and a 
       ruler, when $p$ is a Fermat prime number.
    }
    \label{L-FermatNumbers02}
    \index{Prime number!Fermat} \index{Fermat!prime number}       
    As mentioned \hyperlink{FermatNumbers01}{above} in 
    chapter~\ref{FermatNumbers01}, Fermat wrote to Mersenne regarding 
    his assumption, that all numbers of this type are primes.
    This assumption was disproved by Euler (1732). The prime $641$ divides $f(5)$\footnote{%
    Surprisingly this number can easily be found by using Fermat's theorem (see e.g. \cite[p. 176]{Scheid1994})
    }. 
%    Surprisingly he would have been able obtain a positive result using
%    the negative prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
f(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
f(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
f(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
f(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
f(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = \mbox{65,537} &  \mapsto {\rm ~prime}  \\
f(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = \mbox{4,294,967,297} = 641 \cdot \mbox{6,700,417} &  \mapsto {\rm ~NOT~prime} ! \\
f(6) = 2^{2^6} + 1  = 2^{64} + 1 &  = \mbox{18,446,744,073,709,551,617} \\
                                 &  = \mbox{274,177} \cdot \mbox{67,280,421,310,721} & \mapsto {\rm ~NOT~prime} !\\
f(7) = 2^{2^7} + 1  = 2^{128} + 1 & = \mbox{(see page~\pageref{F7Morrison})}  
																 & \mapsto {\rm ~NOT~prime} !

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{generalizedFermatprimes}{}
\subsection[Generalized Fermat numbers $f(b,n) = b^{2^n} + 1$]
              {Generalized Fermat numbers\footnotemark~$f(b,n) = b^{2^n} + 1$}
    \footnotetext{%
      The base of this power is no longer restricted to 2.
      Even more generic would be:  $f(b,c,n) = b^{c^n} \pm 1$
    }
\label{generalized-fermat}
\index{Fermat!number!generalized} 
    Generalized Fermat numbers are more numerous than Mersenne numbers of a
    equal size and many of them are waiting to be discovered to fill the
    big gaps between the Mersenne primes already found or still undiscovered.
    Progress in number theory made it possible that numbers, where the
    representation is not limited to the base 2, can be tested at almost
    the same speed than a Mersenne number.

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection{Carmichael numbers\index{Number!Carmichael}}
As mentioned \hyperlink{HT-Carmichael-number01}{above} in
chapter~\ref{L-Carmichael-number01} not all Carmichael numbers are prime.

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection{Pseudo prime numbers%
\index{Prime number!pseudo prime}\index{Number!pseudo prime}%
}
See \hyperlink{HT-Pseudoprimenumber01}{above} in
chapter~\ref{L-Pseudoprimenumber01}.

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection{Strong pseudo prime numbers%
\index{Prime number!strong pseudo prime}\index{Number!strong pseudo prime}%
}
See \hyperlink{HT-Strongpseudoprimenumber01}{above} in
chapter~\ref{L-Strongpseudoprimenumber01}.

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection
    [Idea based on Euclid's proof $p_1 \cdot p_2 \cdots p_n +1$]
    {Idea based on Euclid's proof $p_1 \cdot p_2 \cdots p_n +1$}
This idea is based on \hyperlink{euclid}{Euclid's proof} that there are
infinite many prime numbers.
$$
\begin{array}{lll}
2{\cdot}3 +1 &      = 7 &          \mapsto {\rm ~prime} \\
2{\cdot}3{\cdot}5 +1 &      = 31    &      \mapsto {\rm ~prime} \\
2{\cdot}3{\cdot}5{\cdot}7 +1 &      = 211   &      \mapsto {\rm ~prime} \\
2{\cdot}3{\cdots}11 +1 &        = 2,311  &      \mapsto {\rm ~prime} \\
2\cdot3 \cdots 13 +1 &  = 59 \cdot 509 &    \mapsto {\rm ~NOT~prime} ! \\
2\cdot3 \cdots 17 +1 &  = 19 \cdot 97 \cdot 277 &   \mapsto {\rm ~NOT~prime} !
\\
\end{array}
$$

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection{As above but $-1$ except $+1$: $p_1 \cdot p_2 \cdots p_n -1$}
$$
\begin{array}{lll}
2\cdot 3 -1     &   = 5 &   \mapsto {\rm ~prime} \\
2\cdot 3 \cdot  5  -1   &   = 29 &  \mapsto {\rm ~prime} \\
2\cdot 3 \cdots 7  -1   &   = 11 \cdot 19 & \mapsto {\rm ~NOT~prime} ! \\
2\cdot 3 \cdots 11 -1   &   = 2,309 &    \mapsto {\rm ~prime} \\
2\cdot 3 \cdots 13 -1   &   = 30,029 &    \mapsto {\rm ~prime} \\
2\cdot 3 \cdots 17 -1    &  = 61 \cdot 8,369 &   \mapsto {\rm ~NOT~prime!}
\end{array} 
$$

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection[Euclidean numbers $e_n = e_0 \cdot e_1 \cdots e_{n-1} + 1$]
              {Euclidean numbers $e_n = e_0 \cdot e_1 \cdots e_{n-1} + 1$
	       with $ n \geq 1 $ and $ e_0 := 1 $}
%	       with $n$ greater than or equal to $1$ and $e_0 := 1$}
    \index{Euclidean number}
    $e_{n-1}$ is not the $(n-1)$th prime number, but the number 
    previously found here.
    Unfortunately this formula is not open but recursive.
    The sequence starts with

@@

% --------------------------------------------------------------------------
% \pagebreak %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vskip +10 pt
\subsection{$f(n) = n^2 + n + 41$}
\label{L-Polynomfunktion01-41}
   This sequence starts off very {\em promisingly},  
   but is far from being a proof.

@@

% --------------------------------------------------------------------------
\pagebreak %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \vskip +10 pt
\subsection{$f(n) = n^2 - 79 \cdot n + 1,601$}
\label{L-Polynomfunktion02-1601}   \label{SrcPrimArith1}
   This function\footnote{%
   See \hyperlink{PrimArith1}{Appendix A of this chapter} for the source code
   to compute the table using Pari-GP.\index{Pari-GP}
}
   delivers prime numbers for all values from $n=0$ to $n=79$.
Unfortunately $f(80) = 1,681 = 11 \cdot 151$ is not a prime number. To this
date, no function has been found that delivers more prime numbers in a row. On
the other hand, each prime occurs twice (first in the decreasing then in the
increasing sequence), which means that the algorithm delivers a total of 40
different prime values (these are the same ones as delivered by the function
in chapter~\ref{L-Polynomfunktion01-41})\footnote{%
   Another quadratic polynom, which delivers these primes, is: 
   $f(n) = n^2 - 9 \cdot n + 61$.\\
   Among the first 1000 sequence elements more than 50\% are prime
   (See \hyperlink{PrimArith2}{Appendix A of this chapter}).}.
$$
\begin{array}{|ll||ll|}
\hline
f(0)  = 1.601 & \mapsto {\rm ~prim}   &   f(26) =   223 & \mapsto {\rm ~prim} \\
f(1)  = 1.523 & \mapsto {\rm ~prim}   &   f(27) =   197 & \mapsto {\rm ~prim} \\
f(2)  = 1.447 & \mapsto {\rm ~prim}   &   f(28) =   173 & \mapsto {\rm ~prim} \\
f(3)  = 1.373 & \mapsto {\rm ~prim}   &   f(29) =   151 & \mapsto {\rm ~prim} \\
f(4)  = 1.301 & \mapsto {\rm ~prim}   &   f(30) =   131 & \mapsto {\rm ~prim} \\
f(5)  = 1.231 & \mapsto {\rm ~prim}   &   f(31) =   113 & \mapsto {\rm ~prim} \\
f(6)  = 1.163 & \mapsto {\rm ~prim}   &   f(32) =    97 & \mapsto {\rm ~prim} \\
f(7)  = 1.097 & \mapsto {\rm ~prim}   &   f(33) =    83 & \mapsto {\rm ~prim} \\
f(8)  = 1.033 & \mapsto {\rm ~prim}   &   f(34) =    71 & \mapsto {\rm ~prim} \\
f(9)  =   971 & \mapsto {\rm ~prim}   &   f(35) =    61 & \mapsto {\rm ~prim} \\
f(10) =   911 & \mapsto {\rm ~prim}   &   f(36) =    53 & \mapsto {\rm ~prim} \\
f(11) =   853 & \mapsto {\rm ~prim}   &   f(37) =    47 & \mapsto {\rm ~prim} \\
f(12) =   797 & \mapsto {\rm ~prim}   &   f(38) =    43 & \mapsto {\rm ~prim} \\
f(13) =   743 & \mapsto {\rm ~prim}   &   f(39) =    41 & \mapsto {\rm ~prim} \\
f(14) =   691 & \mapsto {\rm ~prim}   &   f(40) =    41 & \mapsto {\rm ~prim} \\              
f(15) =   641 & \mapsto {\rm ~prim}   &   f(41) =    43 & \mapsto {\rm ~prim} \\              
f(16) =   593 & \mapsto {\rm ~prim}   &   f(42) =    47 & \mapsto {\rm ~prim} \\              
f(17) =   547 & \mapsto {\rm ~prim}   &   f(43) =    53 & \mapsto {\rm ~prim} \\              
f(18) =   503 & \mapsto {\rm ~prim}   &   \cdots  &  \\                                       
f(19) =   461 & \mapsto {\rm ~prim}   &   f(77) = 1.447 & \mapsto {\rm ~prim} \\             
f(20) =   421 & \mapsto {\rm ~prim}   &   f(78) = 1.523 & \mapsto {\rm ~prim} \\             
f(21) =   383 & \mapsto {\rm ~prim}   &   f(79) = 1.601 & \mapsto {\rm ~prim} \\             
f(22) =   347 & \mapsto {\rm ~prim}   &   f(80) = 41 \cdot 41 & \mapsto {\rm ~NOT~prim!} \\
f(21) =   383 & \mapsto {\rm ~prim}   &   f(81) = 41 \cdot 43 & \mapsto {\rm ~NOT~prim!} \\ 
f(22) =   347 & \mapsto {\rm ~prim}   &   f(82) = 1.847 & \mapsto {\rm ~prim} \\             
f(23) =   313 & \mapsto {\rm ~prim}   &   f(83) = 1.933 & \mapsto {\rm ~prim} \\              
f(24) =   281 & \mapsto {\rm ~prim}   &   f(84) = 43 \cdot 47 &  \mapsto {\rm ~NOT~prim!} \\
f(25) =   251 & \mapsto {\rm ~prim}   &   & \\
\hline
\end{array} 
$$

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection[Polynomial functions 
    $f(x) = a_n x^n + a_{n-1}x^{n-1} + \cdots + a_1 x^1 + a_0$]
    {Polynomial functions\index{Polynomial} 
    $f(x) = a_n x^n + a_{n-1}x^{n-1} + \cdots + a_1 x^1 + a_0$ ~ 
    ($a_i$ in ${\mathbb Z}$, $n \geq 1$)}\index{Polynomial}
